---
title: "Longitudinal vs Summary"
author: "Matt Shotwell"
date: "Original: 2022-12-13; Updated: `r date()`"
output: 
  html_document:
    code_folding: true
---

## Introduction

### VIOLET

  - Info: https://clinicaltrials.gov/ct2/show/NCT03096314
  - Paper: https://www.nejm.org/doi/full/10.1056/NEJMoa1911124
  - Design: RCT
  - Population: Vitamin D deficient patients at high risk for Acute Respiratory Distress Syndrome (ARDS) and mortality
  - Intervention: Vitamin D vs. placebo 
  - Outcome: 4 level ordinal measured daily for 28 days

```{r packages, message=FALSE}
library('Hmisc')
library('geepack')
library('dplyr')
```

```{r}
## 500 simulated participants from VIOLET
getHdata(simlongord500)

## 250000 simulated participants from VIOLET
getHdata(simlongord)

## convert treatment variable to factor
simlongord500 <- simlongord500 %>%
  mutate(tx = factor(tx, levels=c(0,1), labels=c('control','treatment')))

## separate 'baseline' data
simlongord500_base <- simlongord500 %>%
  filter(time == 1) %>%
  select(id, age, sofa, tx)
```

```{r}
## compute data set for hospital-free days at day 28
simlongord500_hfd28 <- simlongord500 %>%
  group_by(id) %>%
  summarise(hfd28 = ifelse(any(y == 'Dead'), -1, sum(y == 'Home'))) %>%
  left_join(simlongord500_base, by='id')

## fill in the 'missing' days after death so that each participant has 
## a full 28 days worth of data
simlongord500_comp <- simlongord500 %>%
  select(-gap, -age, -sofa, -tx) %>%
  mutate_at(c('y','yprev'), ~as.character(.)) %>%
  full_join(expand.grid(
    id = simlongord500_base$id,
    time = 1L:28L), by=c('id', 'time')) %>%
  left_join(simlongord500_base, by='id') %>%
  mutate_at(c('y','yprev'), ~ifelse(is.na(.), 'Dead', .)) %>%
  mutate_at(c('y','yprev'), ~factor(., levels=c("Dead", "Vent/ARDS", "In Hospital/Facility", "Home"))) %>%
  arrange(id, time)


```

## Figures

### First 14 days of 4 level outcome
```{r}
propsPO(y ~ tx + time, data=simlongord500_comp %>% filter(time <= 14), nrow=1) +
  theme(legend.position='bottom', 
        axis.text.x=element_text(angle=90, hjust=1),
        axis.title.x = element_blank(),
        legend.title = element_blank())
```

## Questions

0. Does non-proportionality in time matter?
  a. Bias and variance of estimates. 
  b. Type-I error.
1. What do we lose in terms of efficiency by using a summary measure
2. If there is informative missingness, which approach is less biased? Scenarios:
  a. Participants that are discharged are more likely to be lost to follow-up
  b. Participants that are discharged are more likely to be lost to follow-up and are subsequently MORE likely to die
  c. Participants that are discharged are more likely to be lost to follow-up and are subsequently LESS likely to die

How to amplify treatment effect: stratify people by risk; sort according to good outcome; top half go to treatment group with probability A > 0.5; bottom half go to control with probability A; roughly half treatment and half placebo; A parameterizes treatment effect where A = 0.5 is null and A=1 is maximal 

0. function to impose treatment effect on 'population'
1. function to simulate study (resample)
1. function to impose missingness (after re-sample)
2. function to compute hfds
3. function to analyze hfds
4. function to analyze longord data
5. functions to summarize results

```{r}

## check for expected data
## dat - data frame; VIOLET simulated data
check_data <- function(dat) {
  ## required data fields
  rfld <- c("id", "time", "y", "age", "sofa", "tx")
  
  ## expected outcome levels
  elev <-  c("Dead", "Vent/ARDS", "In Hospital/Facility", "Home")
  
  ## confirm expected outcome data
  if(!is.data.frame(dat) ||               ## check for data frame
     !all(rfld %in% names(dat)) ||        ## check for all required fields
     !identical(levels(dat$y), elev))     ## check expected outcome levels
    stop('unexpected data')
}

## score participants based on longitudinal outcome
## dat - data frame; VIOLET simulated data
score_sum <- function(dat) {
  check_data(dat)
  
  ## compute score for each participant; sum longitudinally as follows:
  ## "Dead" = 0, "Vent/ARDS" = 1, "In Hospital/Facility" = 2, "Home" = 3
  scr <- dat %>% 
    group_by(id) %>%
    summarise(y_score = sum(unclass(y)-1))
  
  return(scr)
}

order_test <- function(...) {
  browser()
  order(...)
}

## impose treatment effect on 'population'
## dat - data frame; VIOLET simulated data
## prb - probability of assignment to treatment for 'best' half of 
##       participants in each stratum according to score
## sfn - score function 
impose_treatment_effect <- function(dat, prb = 0.5, sfn = score_sum) {
  check_data(dat)

  ## reallocate participants to treatment and control to impose
  ## a treatment effect; parameterized by 'prb'
  bse <- dat %>% 
    ## filter to first time point
    filter(time == 1) %>%
    ## compute and merge outcome scores
    left_join(sfn(dat), by='id') %>%
    ## group age by quartiles
    mutate(age_group = cut(age, breaks=c(-Inf,quantile(age, 1:3/4),Inf))) %>%
    ## group sofa by quartiles
    mutate(sofa_group = cut(sofa, breaks=c(-Inf,quantile(sofa, 1:3/4),Inf))) %>%
    ## stratify by age and sofa groups
    group_by(age_group, sofa_group) %>%
    ## compute normalized rank by score
    mutate(norm_rank = rank(y_score, ties.method='first')/n()) %>%
    ## compute probability of assignment to treatment
    mutate(prob_tret = ifelse(norm_rank >= 0.5, prb, 1-prb)) %>%
    ungroup() %>%
    ## 're-randomize' treatment assignment
    mutate(tx_bin = rbinom(n=n(), size=1, prob=prob_tret)) %>%
    ## format treatment assignment as factor
    mutate(tx = factor(tx_bin, levels=0:1, labels=c('control','treatment')))

  ## merge new treatment assignments with original data
  dat <- dat %>%
    select(-tx) %>%
    left_join(bse %>% select(id, tx), by='id')

  return(dat)
}

## ensure that each participant has 28 rows; that is, for people who died
## before day 28, complete the additional rows to day 28
## dat - data frame; VIOLET simulated data
complete_28day_fu <- function(dat) {
  check_data(dat)

  ## create a 'baseline' data set
  bse <- dat %>% 
    ## filter to first time point
    filter(time == 1) %>% 
    ## select covariates and treatment assignment
    select(id, gap, age, sofa, tx)

  out <- dat %>%
    ## remove items to merge back later
    select(-gap, -age, -sofa, -tx) %>%
    ## factor -> character
    mutate_at(c('y','yprev'), ~as.character(.)) %>%
    ## fill in missing time (days)
    full_join(expand.grid(
      id = bse$id,
      time = 1L:28L), by=c('id', 'time')) %>%
    ## fill in removed items
    left_join(bse, by='id') %>%
    ## all missing values of y and yprev should be 'Dead'
    mutate_at(c('y','yprev'), ~ifelse(is.na(.), 'Dead', .)) %>%
    ## character -> factor
    mutate_at(c('y','yprev'), ~factor(., levels=c("Dead", "Vent/ARDS", "In Hospital/Facility", "Home"))) %>%
    ## sort by id and time (days)
    arrange(id, time)

  return(out)
}

## impose missing values
## no missing until after hospital discharge
## P(D|M) = P(M|D)P(D)/P(M)
## dat - data frame; VIOLET simulated data
## pma - probability of becoming lost to follow-up after hospital discharge for 
##       people who subsequently remained alive through day 28
## pmd - probability of becoming lost to follow-up after hospital discharge for 
##       people who subsequently remained died on or before day 28
impose_missing <- function(dat, pma=) {
  check_data(dat)
}

## code from Frank's website that shows dead after discharge
# ssamp <- sample(unique(d$id), 65, FALSE)
# dr        <- subset(d, id %in% ssamp)
# dr        <- subset(dr, day <= ddeath)
# dr$id     <- as.integer(as.factor(dr$id))
# dr$status <- factor(dr$status, levels=rev(levels(dr$status)))
# dr$day    <- dr$day - 1
# multEventChart(status ~ day + id, data=dr,
#                absorb='Dead', sortbylast = TRUE) +
#   theme_classic() +
#   theme(legend.position='bottom')

```

```{r}
dat_sim <- simlongord500 %>%
  impose_treatment_effect(0.6) %>%
  complete_28day_fu()

propsPO(y ~ tx + time, data=dat_sim %>% filter(time <= 14), nrow=1) +
  theme(legend.position='bottom', 
        axis.text.x=element_text(angle=90, hjust=1),
        axis.title.x = element_blank(),
        legend.title = element_blank())
```

## How often did people die after initial hospital discharge
```{r}
simlongord500 %>%
  group_by(id) %>%
  summarise(died_after_discharge = (function(time, y) {
    mh <- match('Home', y)
    md <- match('Dead', y)
    
    if(!is.na(mh) && !is.na(md))
      return(TRUE)
    
    return(FALSE)
  })(time, y)) %>%
  pull(died_after_discharge) %>%
  table(useNA='always')

simlongord %>%
  group_by(id) %>%
  summarise(died_after_discharge = (function(time, y) {
    mh <- match('Home', y)
    md <- match('Dead', y)
    
    if(!is.na(mh) && !is.na(md))
      return(TRUE)
    
    return(FALSE)
  })(time, y)) %>%
  pull(died_after_discharge) %>%
  table(useNA='always')
```